---
title: "Confidence Intervals"
format: html
editor: 
  markdown: 
    wrap: 72
---

## Confidence Intervals

What is a confidence interval (CI)?

The standard error tells us how much our statistic (like an average)
might change if we took multiple samples. But we can take this one step
further and create a confidence interval (CI), which gives us a range
where we expect the true value to fall.

Think of it like this:

Imagine you want to estimate the average water-holding capacity of
reservoirs in Tikal. Since measuring every reservoir precisely is
impossible, you take a sample of 10 reservoirs, calculate their volumes,
and find the average volume is 50,000 cubic meters.

However, you know that if you sampled another 10 reservoirs, the average
might be slightly different. The standard error helps you estimate how
much the average volume might vary if you took different samples.

Once you have the standard error, you can calculate a confidence
interval (CI), which gives you a range where the true average reservoir
volume likely falls.

### Example

Estimating the Average Reservoir Volume

Step 1: Calculate the Sample Average You measure 10 reservoirs and find
the average volume is 50,000 cubic meters.

Step 2: Estimate the Standard Error The standard error tells you how
much the average might vary if you took another sample. Let’s say the
standard error is 3,000 cubic meters.

Step 3: Create a 95% Confidence Interval A 95% confidence interval is
usually calculated as:

\$\$Sample Average ± ( Standard Error × 2 )
Sample Average±(Standard Error×2)\$\$

For your reservoir data:

50 , 000 ± ( 3 , 000 × 2 ) = 50 , 000 ± 6 , 000
50,000±(3,000×2)=50,000±6,000

So, the 95% confidence interval is 44,000 to 56,000 cubic meters.

What This Means for Your Research You are 95% confident that the true
average reservoir volume in Tikal is between 44,000 and 56,000 cubic
meters. If the confidence interval was wide (e.g., 30,000 to 70,000
cubic meters), it would mean your estimate is less precise because the
standard error is high. If the confidence interval was narrow (e.g.,
48,000 to 52,000 cubic meters), it would mean you have a high level of
precision in your estimate.

```{r}
n <- 10000
mu <- 25
sigma <- 10
set.seed(1)
v <- rnorm(n, mu, sigma)
s <- sample(v, size = 40, replace = FALSE)
(m <- mean(s))
```

```{r}
(se <- sigma/sqrt(length(s)))  # if population standard deviation is known
```

```{r}
(sd <- sd(s))
```

```{r}
(se <- sd(s)/sqrt(length(s)))  # if population standard deviation is unknown
```

```{r}
# alternatively, we could use `sciplot::se(s)`
(se <- sciplot::se(s))
```

To calculate the 95% CI

```{r}
percent_ci <- 95
alpha <- 1 - percent_ci/100  # alpha = 0.05
lower <- m + qnorm(alpha/2) * se
# where qnorm(alpha /2) is the 2.5% quantile of the standard normal
# distribution
upper <- m + qnorm(1 - alpha/2) * se
# where qnorm(1 - alpha / 2) is the 97.5% quantile of the standard normal
# distribution
(ci <- c(lower, upper))
```

### Beta Distribution

```{r}
library(mosaic)
```

```{r}
plotDist("beta", shape1 = 0.3 , shape2 = 4)
reps = 1000
s <- do(reps) * mean(rbeta(n=100, shape1 = .3, shape2 = 4))
histogram(s$mean)
```

rnorm() and rbeta() draw random numbers from different distributions
rnorm() = normal distribution random rbeta() - beta distribution random

```{r}
x <- rbeta(n = 1000, .3, 4)
se <- sd(x)/sqrt(length(x))
se
```

```{r}
histogram(x)
```

```{r}
reps <- 500

s <- do(reps) * mean(rbeta(n=100, .3, 4)) #500 means of 100 random samples
s
```

```{r}
sd(s$mean)
```

```{r}
histogram(s$mean)
```

## CRITICAL VALUES

We will often use q\_() function to get the value of X associated with
particular quantiles of a distribution, whether that distribution is
theoritcal or empirical

```{r}
c <- qnorm(c(0.025, 0.50, 0.975), mean = 0, sd = 1)
c
```

```{r}
q <- qbeta(p = c(.025, .975), shape1 = 2, shape2 = 4)
q
```

Example

```{r}
plotDist("norm", mean = 10, sd = 2)
c <- qnorm(p = c(0.025, 0.25, 0.50, 0.75, 0.975), mean = 10, sd = 2)
c #displays values that fall at 2.5%, 25%, 50%, 75%, and 97.5%, results are our critical values
```

What is the theoritical 95% CI around the estimate of the mean of the
following vector

```{r}
x <- c(2.9, 4.8, 8.9, -3.2, 9.1, -2.5, -0.9, -0.1, 2.8, -1.7) #random ten numbers from a normal distribution
m <- mean(x)
m
se <- sd(x)/sqrt(length(x)) #estimate of uncertainty of estimate of the mean based on my ten
se
ci <- m + qnorm(c(0.025, 0.975)) * se #
ci #95% of the time we would expect our values to fall within this range
```

```{r}
percent_ci <- 0.95
alpha <- 1- percent_ci/100
alpha
ci <- m + qnorm(c(alpha/2, (1-(alpha/2)))) * se
ci
```

## T-Distribution for smaller \<30 population sizes

```{r}
mean(x) + c(qt(0.025, df = length(x)-1), qt(0.975, df = length(x)-1))*sd(x)/sqrt(length(x))
```

Instead of assuming its normal, resample, use data from a single sample
and we use monte carlo to approximate a sampling distribution even
without access to the population from which samples are drawn

```{r}
x <- c(2.9, 4.8, 8.9, -3.2, 9.1, -2.5, -0.9, -0.1, 2.8, -1.7)
n_boot <- 1000 #elements long 10000
n<- length(x)
boot <- vector(length = n_boot) #set up a dummy variable to hold our simulations
n <- length(x) #bootstrap sample size will be the same length as our sample data

for (i in 1:n_boot){
  boot[[i]] <- mean(sample(x, n, replace = TRUE))
}
ci<- quantile(boot, probs = c(0.025, 0.975))
```

```{r}
boot
```
