---
title: "Class Challenge"
format: html
---

## A Two Sample T-test

```{r}
library(readr)
f <- "https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/tbs-2006-2008-ranges.csv"
d <- read_csv(f, col_names = TRUE) 
```

```{r}
#determine mean, SD, and SE, in Kernel 95 home range size for each sex
kernel95_stats <- d |>
  group_by(sex) |>
  summarize(
    
    #calculate mean
    mean = mean(kernel95),
    
    #calculate SD
    sd = sd(kernel95),
    
    #calculate se
    se = sd(kernel95)/sqrt(length(kernel95))
  )
```

```{r}
p <- ggplot(data = d, mapping = aes(x = sex, y = kernel95)) +
  geom_boxplot() +
  geom_jitter(color="black", size=0.4, alpha=0.9)

p
```

For each sex, generate a bootstrap distribution for the mean of home range size, where you resample the data within each sex 10,000 times

```{r}
library(mosaic)

m$sex <- "males"
f$sex <- "females"

n_boot <- 10000

boot_m <- do(n_boot) * mean(sample(M$kernel95, length(M$kernel95), replace = TRUE))
```

Inferential analysis
Is there a significant difference in mean home range size of males and females?
What is H0 and Ha?

Two samples t-test assuming roughly equal variance in the two groups
  What is the test statistic?
  
```{r}
t_stat <- (mean_m$mean mean_f$mean) - 0

s2 <- ((nrow(males) - 1) * sd(males$kernel95)^2 + (nrow(females)-1) * sd(females$kernel95)^2)
```

```{r}
t.test(x = m$kernel95_stats, y = f$kernel95_stats, alternative = "two.sided")
```

```{r}
d2 <- data.frame(
  sex = d$sex,
  kernel95 = d$kernel95
)

d2
```

```{r}
library(dplyr)
d <- d |>
  select(id, sex, kernel95)

summary <- d |>
  group_by(sex) |>
  summarize(mean = mean(kernel95))

obs <- filter(summary, sex == "F") |> pull(mean) -
        filter(summary, sex == "M") |> pull(mean)

obs
```
generating null distribution from our data
```{r}
reps <- 10000
perm <- vector()

#shuffling up sex vector
for (i in 1: reps){
  temp <- d
  temp$sex <-sample(temp$sex)
  summary <- temp |>
    group_by(sex) |>
    summarize(mean = mean(kernel95))
  perm[[i]] <- filter(summary, sex == "F") |> pull(mean) - 
    filter(summary, sex == "M") |> pull(mean)
}

hist(perm)
```

```{r}
p <- sum(perm < -1 * abs(obs) | perm > abs(obs))/reps
p
```

```{r}
#install.packages("infer")
library(infer)
```

```{r}
d <- d |> specify(formula = kernel95 ~ sex)
d <- d |> hypothesize(null = "independence")
perm <- d |> generate(reps = 10000, type = "permute")
perm <- perm |> calculate(stat = "diff in means", order = c("M", "F"))
perm
visualize(perm, bins = 20)
```

read module 16 and exercise 6