---
title: "Statistics"
format: html
---

## NORMAL Distribution

```{r}
install.packages("mosaic")
```

```{r}
library(mosaic)
```

```{r}
mu <- 10 # for 'mean' parameter
sigma <- 2 # for 'sd' parameter
plotDist("norm", mean = mu, ds = sigma, xlab = "x", ylab = "Frequency")
```

```{r}
s1 <- rnorm(n=10, mean = 10, sd = 2) #draws different set of random numbers from this set each time, so our mean will always be slighlty different
mean(s1)
```

```{r}
s2 <- rnorm(n = 1000, mean = 10, sd = 2)
mean(s2)
sd(s2)
```

### SAMPLING Distribution

Different from a normal distribution, distribution of a statistic. So for instance the mean of s1 will be different each time because each time it is taking a random set of 10 numbers.

```{r}
reps <- 500 #generate 500 replication

samp_dist_mean <-
  do(reps) * mean(rnorm(n = 10, mean = 10, sd = 2)) #repeated sampling out of normal distribution, we are doing this 500 times
#the thing i am doing 500 times is taking the mean of a normal distribution drawn from a sample of 10 that has a mean of 10 and a sd of 2
str(samp_dist_mean)
# generates a sampling distribution for the mean of our sample
samp_dist_median <-
  do(reps) * median(rnorm(n = 10, mean = 10, sd = 2))
# generates a sampling distribution for the median of our sample
```

```{r}
histogram(samp_dist_mean$mean)
```

```{r}
histogram(samp_dist_median$median)
```

```{r}
mean(samp_dist_mean$mean)
```

## RELIABILITY

How far off is a statistic that we calculate based on a sampling distribution likeley to be from the true population value of the parameter interest

standard error - measure of reliability/unreliability standard error (SE) = square root of the variance of the sampling distribution = standard deviation of a sampling distribution

standard deviation of the mean variable of that distribution sd \<- applies to a sample se \<- measure of uncertainity in a sampling distribution, sd of sampling distribution

```{r}
se_mean <- sd(samp_dist_mean$mean) #How uncertain is estimate of the mean
se_median <- sd(samp_dist_median$median) #How uncertain is the estimate of the median
```

```{r}
sd(samp_dist_mean$mean)
```

## Confidence Intervals

The SE can be used to derive another measure of uncertainty in a statistic's value: the confidence interval or the CI

The CI is another way of describing a statistic's sampling distribution, and it plays a central role in basic inferential statistics.

a 95% CI for instance, if i were to repeat the sampling again, 95% of the time the CI should fall within that CI that I estimated within my first sample.

```{r}
challenge <- rnorm(n = 1000, mean = 2, sd = 4)
mean(challenge)
sd(challenge) 
```

Standard Error of the mean based on this sample

SE = sd(x)/sqrt(n)

```{r}
se <- sd(challenge)/sqrt(1000)
```

```{r}
reps <- 1000 #generate 500 replication

s_mean <-
  do(reps) * mean(rnorm(n = 100, mean = 2, sd = 4)) #repeated sampling out of normal distribution, we are doing this 500 times
#the thing i am doing 500 times is taking the mean of a normal distribution drawn from a sample of 10 that has a mean of 10 and a sd of 2
str(samp_dist_mean)
# generates a sampling distribution for the mean of our sample
s_median <-
  do(reps) * median(rnorm(n = 100, mean = 2, sd = 4))
# generates a sampling distribution for the median of our sample
```

```{r}
se_mean <- sd(s_mean$mean) #How uncertain is estimate of the mean
se_median <- sd(s_median$median) #How uncertain is the estimate of the median
```

```{r}
histogram(s_mean$mean)
```

Beta Distribution

```{r}
plotDist("beta", shape1 = 0.3 , shape2 = 4)
reps = 1000
s <- do(reps) * mean(rbeta(n=100, shape1 = .3, shape2 = 4))
histogram(s$mean)
```
